Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Typos found by Andrea, beginning of sec 3

Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.  For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.  The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and show be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check can be hazardous, especially
when the equality-check is type directed.

Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.  The most interesting rules are the $\eta$-rules for unit and function type and the congruence rule for type application:
\begin{gather*}
\renewcommand{\jchk}{:}
\renewcommand{\jinf}{:}
  \ru{}{\Gamma \der t = t' \jchk \One}
\qquad
  \ru{\Gamma, x \of A \der t\,x = t'\,x \jchk B
    }{\Gamma \der t = t' \jchk A \to B}
\qquad
  \ru{\Gamma \der t = t' \jinf \forall X. B
    }{\Gamma \der t\,A = t'\,A \jinf B[A/X]}
\end{gather*}
The Curry-style version replaces the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.  However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.  How would we adopt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?  The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.  For instance, the algorithm would yield $t\,\One\,x = t\,(A \to A)\,y$ even for $x \not= y$.
We could also consider a heterogeneous check $(\Gamma \der t : A) = (\Gamma' \der t' : A')$ where each term is paired with its own type and context, but this leaves us with the dilemma of explaining the meaning of this judgement when $A$ and $A'$ are incompatible.

@@

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step what $\eta$-expansion concerns, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

@@

Syntax (de Bruijn terms).
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_1} t_2 \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size
   & \mbox{contexts} \\
\Subst & \ni & \eta,\rho,\sigma,\tau,\xi & ::= & \sempty \mid (\sigma,t)
  & \mbox{substitutions} \\
\end{array}
\]

%We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
We write $t \sigma$ for the application of substitution $\sigma$ to term $t$ which is defined as usual.  We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.  We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$. In general, we refer to liftings by letter $\xi$.
The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.
Substitution composition $\sigma\tau$ is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
\begin{gather*}
  t\sid \equiv t
\quad
  (t\sigma)\tau \equiv t(\sigma\tau)
\qquad
  \sigma \sid \equiv \sigma
\quad
  \sid \tau \equiv \tau
\quad
  (\rho\sigma)\tau \equiv \rho(\sigma\tau)
\\
  \ind 0 (\sigma,t) \equiv t
\quad
  \slift(\sigma,t) \equiv \sigma
\qquad
  [t]\sigma \equiv (\sigma, t\sigma)
\quad
  \slift[t] \equiv \sid
%  (\sigma\slift)[t] \equiv \sigma
\end{gather*}

We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default.  We write $\forallT T$ for $\erpiT \Size T$.
For concrete function types, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
We may also use names when we construct concrete contexts, for instance, if $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
$\Gamma.\Size.\Nat \ind 0$.

>>>>

Higher-rank size quantification takes termination checking to the next level; it is necessary for abstraction and generic programming.  For instance, it allows us to write a generic tree traversal which applies a user-given preprocessor on subtrees before recursively descending into these trees.  The condition is that preprocessing does not increase the size of the subtree; otherwise, termination could not be guaranteed.  Concretely, assume a type $\tT\,i$ of trees of size $< i$ with a constructor $\tnode : \forall i.\, \List\,(\tT\,i) \to \tT\,(i+1)$ which takes a finite list of subtrees to form a new tree.
In the following definition of $\ttrav$, the preprocessing $\vpre : \forall i.\, \tT\,i \to \tT\,i$ can be safely applied to input tree $t$ because the type of $\vpre$ bounds the size of $\vpre\,t$ by the size of $t$.
In case $\vpre\,t = \tnode\,\vts$, the trees in the list $\vts$ are still guaranteed to be of strictly smaller size than $t$, thus, the recursive call to $\ttrav$, communicated via the $\tmap$ function for lists, is safe.
\[
\begin{array}{l}
  \ttrav : (\vpre : \forall i.\, \tT i \to \tT i) (\vpost : \tT \infty \to \tT \infty) \to \forall i.\, \tT\,i \to \tT\,\infty
  \\
  \ttrav \; \vpre \; \vpost \; t = \vpost\;(\tcase\;\vpre\,t\;\tof \, \{ \, \tnode\,\vts \to \tnode\,(\tmap \; (\ttrav \, \vpre \, \vpost) \; \vts)\, \})
\end{array}
\]
The display shows the \emph{Curry}-style program as provided by the user, but state-of-the-art type checkers elaborate the program from surface syntax into an internal \emph{Church}-style syntax with explicit type abstractions and type applications.\footnote{Agda, Coq, Idris, and Haskell \cite{sulzmannChakravartyPeythonJonesDonnelly:tldi07} all have Church-style internal languages.}  With implicit type and size applications elaborated, $\ttrav$ would look as follows:
\[
\begin{array}{l}
  \ttrav \; \vpre \; \vpost \; i\; t = \vpost\;(\tcase\;\vpre\,i\,t\;\tof \, \{ \, \tnode\,j\,\vts \to \tnode\,\infty\,(\tmap \; (\tT\,j) \; (\tT\,\infty) \; (\ttrav \, \vpre \, \vpost \, j) \; \vts) \})
\end{array}
\]
Church-style syntax is the basis for all program analyses and transformations to follow and should be considered as the \emph{true} syntax.
However, from a dependent-type perspective, explicit size applications
in terms can be problematic when the type checker compares terms for
equality---which is necessary as types can depend on values.  Inferred
sizes may not be unique, as we have subtyping $\tT\,i \leq \tT\,j$ for
$i \leq j$: we can always weaken an upper bound.  For instance, given
$\vts : \List\,(\tT\,i)$, any of the terms
$\tnode\,i\,\vts$, $\tnode\,(i+1)\,\vts$, $\dots$, $\tnode\,\infty\,\vts$
%$\tnode\,i\,\vts$ or $\tnode\,(i+1)\,\vts$ or $\dots$ or $\tnode\,\infty\,\vts$
has type
$\tT\,\infty$.  Yet semantically, all these trees are equal,
thus, the syntactic equality check should ignore the size argument to
$\tnode$.  Similarly, in the application $\vpre\,i\,t$ the size
argument $i$ should be ignored by the equality check.  Yet
$\vpre\,i : \tT\,i \to \tT\,i$ and $\vpre\,j : \tT\,j \to \tT\,j$ have
different types for $i \not= j$, and moreover these function types are
not in the subtyping relation due to the mixed-variant occurrence of the
size parameter.  It seems that during equality checking we have to
consider terms of different types, at least for a while.  Once we apply
$\vpre\,i$ and $\vpre\,j$ to the same tree $t : \tT\,k$, which
determines $i = j = k$, we are back to safety.  However, allowing
types to differ during an equality check can be hazardous, especially
when the equality-check is type directed.

Consider the analogous situation for the polymorphic lambda calculus System~F, be it the predicative variant or not, extended by a unit type $\One$.
For Church-style, we can give a type-directed $\beta\eta$-equality test which equates all terms at the unit type.  The most interesting rules are the $\eta$-rules for unit and function type and the congruence rule for type application:
\begin{gather*}
\renewcommand{\jchk}{:}
\renewcommand{\jinf}{:}
  \ru{}{\Gamma \der t = t' \jchk \One}
\qquad
  \ru{\Gamma, x \of A \der t\,x = t'\,x \jchk B
    }{\Gamma \der t = t' \jchk A \to B}
\qquad
  \ru{\Gamma \der t = t' \jinf \forall X. B
    }{\Gamma \der t\,A = t'\,A \jinf B[A/X]}
\end{gather*}
The Curry-style version replaces the last conclusion by $\Gamma \der t = t' : B[A/X]$ where the type $A$ to instantiate $X$ has to be guessed.  However, in Curry-style more terms are equated than in Church-style, as for instance the Church-style terms $t\,A\,(\lambda x \of A.\, x)$ and $t\,B\,(\lambda x \of B.\, x)$ map to the same Curry-style term $t\,(\lambda x.\,x)$.  How would we adapt the algorithm for Church-style such that it equates all terms that are equal in Curry-style?  The conclusion of the last rule could be changed to $\Gamma \der t\,A = t'\,A' : B[A/X]$, but then the second term $t'\,A'$ does not have the ascribed type $B[A/X]$, and $\eta$-laws applied to this term would be unsound.  For instance, the algorithm would yield $t\,\One\,x = t\,(A \to A)\,y$ even for $x \not= y$.
We could also consider a heterogeneous check $(\Gamma \der t : A) = (\Gamma' \der t' : A')$ where each term is paired with its own type and context, but this leaves us with the dilemma of explaining the meaning of this judgement when $A$ and $A'$ are incompatible.

@@

The central technical observation is that reflection and reification do not need the precise type $T$, they work the same for any \emph{shape} $S$ of $T$.
We managed, while not introducing a new syntax for shapes, to define a relation $T \shape S$ on type values stating that type $S$ qualifies as shape for type $T$.
Shapes unfold dynamically during reflection and reification.
For example, when reflecting a variable $x$ into the polymorphic function type $\forall i.\, F\,i$ where $F\,i = \Nat\,i \to \Nat\,i$, we obtain
$(\up{\forall i.\/ F\,i}x)\,i = \up{F\,i} (x\,i)$ for size $i$ and
$(\up{\forall i.\/ F\,i}x)\,j = \up{F\,j} (x\,j)$ for size $j$.  The new types $F\,i$ and $F\,j$ we reflect at are no longer equal (and they are not subtypes of each other), but they still have the same shape, $\Nat\_ \to \Nat\_$.  This means they will still move in lock-step in respect to $\eta$-expansion, which is sufficient to prove NbE correct for judgmental equality.
We call the enabling property of $F$ \emph{shape irrelevance}, meaning that for any pair $i$, $j$ of legal arguments, $F\,i$ and $F\,j$ have the same shape.  Whenever we form a irrelevant function type $\forall x \of U.\, T[x]$, we require $T[x]$ to be shape-irrelevant in $x$.  This is the middle ground between ICC$^*$, where no restriction is placed on $T$ but $\eta$ for unit types is out of reach (at least for the moment), and Pfenning's \citeyearpar{pfenning:intextirr} irrelevance modality, adapted to full dependent types by \citet{abelScherer:types10}, which requires $T$ to be irrelevant in $x$ and, thus, has type equality $T[i] = T[j]$.

@@

In this section, we give syntax and \emph{declarative} typing, equality, and subtyping judgements.  The typing relation $\Gamma \der t : T$ will \emph{not} be decidable; instead, we present algorithmic typing $\Gamma \der t \jchk T$ in Section~\ref{sec:tycheck}.  However, equality and subtyping will be decidable for well-formed input, see sections~\ref{sec:sem}--\ref{sec:asub}.

We present our type theory as \emph{pure type system} \cite{barendregt:lambdacube} with extra structure.
\emph{Sorts} $s$ are drawn from an infinite predicative hierarchy of universes $\Set[\ell]$ for $\ell \in \NN$.  Universes provide us with polymorphism and the capability to define types by recursion on values.  Whether we have just two universes $\Set[0]$ and $\Set[1]$ or infinitely many, does not matter for the technical difficulty of the meta theory.  The present setup have the advantage that every sort has again a sort since $\Set[\ell] : \Set[\ell+1]$, thus, we do not have to introduce a separate judgement $\Gamma \der T$ for well-formedness of types, we can define it as $\exists s.\ \Gamma \der T : s$.

\begin{figure}[htbp]
  \centering
\[
\begin{array}{lllrl@{\qquad}l}
\Sort & \ni & s
    & ::= & \Set[\ell] ~(\ell \in \NN)& \mbox{sort (universe)} \\
% \Ann & \ni & \star & ::= & \erased \mid \nonstrict \mid \noterased & \mbox{annotation (irrelevant, shape-irrelevant, relevant)}\\
\Ann & \ni & \star & ::= & \erased \mid \noterased & \mbox{annotation (irrelevant, relevant)}\\
\Exp & \ni & t,u,T,U
    & ::= & w \mid t\,e & \mbox{expressions} \\
\Whnf & \ni & w, W
    & ::= & n \mid s \mid \Size \mid \epiT U T % \mid \piSizeT T
    % \mid \forallT T
    \mid \lambda t% & \mbox{functions} \\ &&
    \mid \Nat a \mid c & \mbox{weak head normal forms} \\ % \mbox{natural numbers} \\
\Data & \ni & c
   & ::= & \zero a \mid \suc a t & \mbox{constructed data} \\
\Ne & \ni & n
    & ::= & \ind i \mid n \, e & \mbox{neutral expressions} \\
\Elim & \ni & e
    & ::= & t \mid a \mid \ann a \mid \case[\ell] T {t_z} {t_s} \mid \fix[\ell] T t & \mbox{eliminations} \\
\SizeExp & \ni & a,b & ::= & \infty \mid o \mid \ind i + o ~(o \in \NN) & \mbox{size expressions} \\
\Cxt & \ni & \Gamma, \Delta & ::= & \cempty \mid \erext[\noterased] \Gamma T \mid \erext \Gamma \Size
   & \mbox{contexts} \\
\Subst & \ni & \eta,\rho,\sigma,\tau,\xi & ::= & \sempty \mid (\sigma,t)
  & \mbox{substitutions} \\
\end{array}
\]
  \caption{Syntax.}
  \label{fig:syntax}
\end{figure}

For the expression syntax (see Fig.~\ref{fig:syntax}), we use de~Bruijn \citeyearpar{deBruijn:nameless} indices $\ind i$ to represent variables.  Binders are lambda abstraction $\lambda t$ and dependent function types $\epiT U T$, which bind the $0$th index in $t$ and $T$, resp.  The notation $\epiT U T$ is an umbrella for three kinds of function types, where $\evar \in \{ \erased, \noterased \}$ is a relevance annotation borrowed from \citet{pfenning:intextirr}: $\erpiT[\noterased] U T$, the ordinary dependent function type, $\erpiT[\noterased] \Size T$ for relevant size quantification, and $\erpiT \Size T$ for irrelevant size quantification.
We omit the ``$\noterased$''-markers from $\Gamma$ and $\Pi$ by default and write $\forallT T$ for $\erpiT \Size T$.
%
Examples for relevant size quantification $\piT \Size T$, in syntax with named variables, are $\Size \to \Set[0]$ or $\funT i \Size \Nat i \to \Set[0]$.  An instance of irrelevant quantification $\forall T$ would be $\forall i.\, \Nat i \to \Nat i$.  Herein, $\Nat i$ denotes the type of natural numbers below $i$. The expression $\Size$ is a possible instance of $U$ in $\epiT U T$, or a possible type of a variable in a typing context $\Gamma$, but not a first-class type, \ie, we cannot construct our own function on sizes.

Canonical natural numbers $c$ are constructed by $\zero a$ and $\suc a t$.  Size expression $a$ is either a constant $o \in \NN$, a variable $\ind i + o$ possibly with increment $o$, or the limit ordinal $\infty$ which stands for $\omega$.  The size expression $s$ in the constructors is a suggestion for the type checker but bears no semantic significance.  For example, in the declarative typing presented here, we can have $\der \zero 5 : \Nat 1$; in the algorithmic typing however, $\der \zero 5 \jchk \Nat 1$ will be an error.

%We write $\sigma$ and $\tau$ for substitutions, which are lists of terms, and
Substitutions $\sigma$ are simply lists of terms which provide one term as replacement for each free de Bruijn index in a term $t$. We write $t \sigma$ for the application of substitution $\sigma$ to term $t$ which is defined as usual.  We write $\slift^k_m$ for the substitution $(\ind{k+m-1},\dots,\ind{k+1},\ind k)$ which applies to a term with $m$ free indices and increases all of them by $k$.  We write $\slift_m$ for the lifting $\slift^1_m$ and $\sid_m$ for the identity substitution $ \slift^0_m$. In general, we refer to liftings by letter $\xi$.
The substitution $[u]_m = (\sid_m, u)$ replaces free index $\ind 0$ by term $u$ and decrements all other free indices by $1$. We drop subscript $m$ when clear from the context.
Substitution composition $\sigma\tau$ is the pointwise application of substitution $\tau$ to the list of terms $\sigma$.  In the proofs to follow, we freely use the following identities:
\begin{gather*}
  t\sid \equiv t
\quad
  (t\sigma)\tau \equiv t(\sigma\tau)
\qquad
  \sigma \sid \equiv \sigma
\quad
  \sid \tau \equiv \tau
\quad
  (\rho\sigma)\tau \equiv \rho(\sigma\tau)
\\
  \ind 0 (\sigma,t) \equiv t
\quad
  \slift(\sigma,t) \equiv \sigma
\qquad
  [t]\sigma \equiv (\sigma, t\sigma)
\quad
  \slift[t] \equiv \sid
%  (\sigma\slift)[t] \equiv \sigma
\end{gather*}

For concrete function types, we may use a named dependent function type notation as syntactic sugar for the corresponding de Bruijn representation.  For instance, $\funT i \Size \Nat i \to \Set[\ell]$ is sugar for $\piT \Size \piT {(\Nat\,\ind 0)} \Set[\ell]$.  We abbreviate this type by \fbox{$\FixK\,\ell$}, and let \fbox{$\FixT\,T$} stand for $\allT i \Size (\funT x {\Nat i} T\, i\, x) \to \funT x {\Nat (i + 1)} T \,(i+1)\, x$. Similar as for $\Pi$, we use named lambda abstraction as sugar for de Bruijn abstraction.  Named abstract takes care of proper lifting of de Bruijn indices, for instance, $\lambda x.\, t x = \lambda. (t \slift)\,\ind 0$ if $t$ is outside the scope of $x$.
We may also use names when we construct concrete contexts, for instance, if $T$ is well-formed in context $\Gamma$, we may write $T\,i\,x$ in context $\Gamma.i\of\Size.x\of\Nat i$ to mean $T\slift^2\,\ind 1\,\ind 0$ in context
$\Gamma.\Size.\Nat \ind 0$.
